# Exercise 0: Appropriately set-up your environment (clear environment, load packages, set working directory if needed, etc.)

```{r setup, include=FALSE}


```

# Introduction

In this lab, we will delve deeper into population dynamics, exploring both density-independent and density-dependent growth, as well as interspecific interactions. You will use R to simulate these ecological processes and interpret the results of your simulations.

# Exercise 1:

Using the `tribbles.rds` dataset provided, write an R script to plot the population size of tribbles over **5 days** using `ggplot2` for plotting and `dplyr` for data wrangling. Plot using the `geom_line()` function.

::: callout-tip
## \`.Rds\` and \`.Rda\` files in R

`.rds` and `.rda` (or `.Rdata`) files are used to store data natively in R.

The main advantage of `.rda` files is that they can contain a variety of data in a single file (objects, plots, functions, packages, etc.), however when loading a `.rda` file you run the risk of over-writing existing objects, plots, etc. with the same name.

`.rds` files can only contain a single data structure. A benefit of this file type is that loading a `.rds` file will not overwrite data structures with the same name.
:::

```{r}

# Load the data
tribbles <- readRDS(file = "tribbles.rds")

# Filter the data


# Plot the filtered data, hint: use xlim(c(1, 5)) to set limits for the x-axis to only show up to day 5



```

### Q1: What general trend do you observe in the growth of the tribble population over this period? Does the population appear to be increasing, decreasing, or remaining stable?

### Q2: Does population growth apper to be discrete or continuous? Why?

### Q3: Does population growth appear to be density-independent? Why or why not? Which equation, if any, do you think might best fit the trend?

### Q4: If you think it may be a density-dependent trend, what would you estimate is the carrying capacity of the tribble population?

### Q5: What are some biological, ecological, or stoichastic explanations that could explain the observed trend?

### Q6: What might be some expected consequences of this trend for the tribble population for the people living on the ship?

### **Step 2: Extract Growth Parameters**

1.  Use **dplyr** to extract the **initial population size** ((N_0)) (*hint, remember `pull()` from the previous assignment*) and **growth rate** ((\lambda)).
2.  We assume the per capita growth rate ((\lambda)) can be estimated as the ratio between consecutive population sizes. (*hint, this is a ratio between N0 and our current day, which is day 5*)

```{r}
# Extract initial population size (N0)


# Estimate lambda using the ratio of consecutive population sizes


# Print estimated values

```

Like we did in lab, use our equation for Q3 and `sapply()` to calculate when the tribble population size for each day until the population exceeds **111,537 tribbles** based on our current assumptions - If you think density-independent, you can use lambda directly in your equation. - If you think logistic growth, remember from lab how to convert lambda to r. We will assume K is equal to the largest population number within the given time span.

```{r}
# Set the value for t


# Use sapply() to calculate population sizes


```

Plot your results like we did in lab:

```{r}

# convert to dataframe

# plot

```

### Q: Does this model match with your plot of the real data above?

### Q: What does this plot imply about the population growth of tribbles on the ship?

------------------------------------------------------------------------

# Exercise 2:

Now plot over 8 days and answer the questions below based on this plot.

```{r}

# Filter the data


# Plot the filtered data. Hint, use xlim(c(1, 8))  


```

### Q1: What general trend do you observe in the growth of the tribble population over this period? Does the population appear to be increasing, decreasing, or remaining stable?

### Q2: Does population growth apper to be discrete or continuous? Why?

### Q3: Does population growth appear to be density-independent? Why or why not? Which equation, if any, do you think might best fit the trend?

### Q4: If you think it may be a density-dependent trend, what would you estimate is the carrying capacity of the tribble population?

### Q5: What are some biological, ecological, or stoichastic explanations that could explain the observed trend?

### Q6: What might be some expected consequences of this trend for the tribble population for the people living on the ship?

### **Step 2: Extract Growth Parameters**

1.  Use **dplyr** to extract the **initial population size** ((N_0)) (*hint, remember `pull()` from the previous assignment*) and **growth rate** ((\lambda)).
2.  We assume the per capita growth rate ((\lambda)) can be estimated as the ratio between consecutive population sizes. (*hint, this is a ratio between N0 and our current day, which is day 8*)

```{r}
# Extract initial population size (N0)


# Estimate lambda using the ratio of consecutive population sizes


# Print estimated values

```

Like we did in lab, use our equation from Q3 and `sapply()` to calculate when the tribble population size for each day until the population exceeds **111,537 tribbles** based on our current assumptions. - If you think density-independent, you can use lambda directly in your equation. - If you think logistic growth, remember from lab how to convert lambda to r. We will assume K is equal to the largest population number within the given time span.

```{r}
# set t


# Use sapply() to calculate population sizes


```

Plot your results like we did in lab:

```{r}

# convert to dataframe

# plot

```

### Q: Does this model match with your plot of the real data above?

### Q: What does this plot imply about the population growth of tribbles on the ship?

------------------------------------------------------------------------

# Exercise 3:

Now plot all time points from the dataset and answer the questions below based on this plot. Use the function `geom_step()` to plot the data.

```{r}

# create df with full data set (do not filter)


# plot



```

### Q: What general trend do you observe in the growth of the tribble population over this period? Does the population appear to be increasing, decreasing, or remaining stable?

### Q: Does population growth apper to be discrete or continuous? Why?

### Q: Does population growth appear to be density-independent? Why or why not? Which equation, if any, do you think might best fit the trend?

### Q: If you think it may be a density-dependent trend, what would you estimate is the carrying capacity of the tribble population?

### Q: What are some biological, ecological, or stoichastic explanations that could explain the observed trend?

### Q: What might be some expected consequences of this trend for the tribble population for the people living on the ship?

------------------------------------------------------------------------

#### Opinion: what was the most useful and least useful aspect of this lab assignment? Why is that the case for you?

``` plaintext
```

# References

#### What outside sources did you use to help you complete this work (including books, forums, LLMs, etc.)? Describe how they were used:

``` plaintext
```
