# Working with real interspecific data

For this assignment, you will search online to find and analyze a dataset other than the one used in lab to see if competition appears to be occurring between 2 species. Try using resources like Dryad, Google Dataset Search, Zenodo among others to find a dataset.

## **Finding a Dataset Tips**

To find a dataset that will work it ideally matches these 3 criteria:

-   At least 2 different species

-   Population counts for each species or each row representing a single individual

-   Population counts at different time points (at least 8 time points)

If you are able to find a good dataset you want to use but there is not a column for population counts, but each *row* is an *individual* (like in a dataset of mice species, each row is an individual mouse), you can still use that dataset by creating a population count column like this:

```{=plaintext}

df_count <- df %>% 
    group_by(species, time) %>% 
    summarize(count = n(), .groups = "drop") %>% # Using summarize() instead of mutate() to make grouping permanent 
    ungroup()
```
If your dataset doesn't give total population at a given time, but instead gives additional individuals observed at each time point, you can use the following to calculate cumulative count:

```{=plaintext}

df_count_cumulative <- df_count %>%
    arrange(species, time) %>% 
    mutate(cumulative_count = cumsum(count))
```
## Exercise 0: Set-up and load data

```{r setup, include=FALSE}



```

## 1: Describe your dataset and how it was collected

```{=plaintext}
```
## 2: Explore your dataset and describe what you see

```{r}


```

## 3: Clean your data using dplyr

```{r}



```

## 4: Plot your data for initial analysis/description of your dataset. What trends do you see?

```{r}



```

## 5: Define the Lotka-Voltera interspecific competition model function

```{r}



```

## 6: Extract the argument values from your dataset necessary as inputs into the Lotka-Voltera interspecific competition model

```{r}

# Times



```

```{r}

# N01, N02



```

```{r}

# r1, r2



```

```{r}

# a11, a22


```

```{r}

# a21, a12


```

## 7: Run your model and plot using `ggplot2`

```{r}

# Run with both species present

# Set parameters


# Set N0


# Run the ODE model


# Convert to dataframe for plotting


# Plot the results


```

```{r}

# Run with one species absent

# Set N0


# Run the ODE model


# Convert to dataframe for plotting


# Plot the results


```

```{r}

# Run with the other species absent

# Set parameters


# Set N0


# Run the ODE model


# Convert to dataframe for plotting


# Plot the results


```

## 8: Interpret your results, what do you observe?

```{=plaintext}

```

## 9: Bonus, plot a zero growth isocline analysis on your dataset:

```{r}


```

### What do you observe? Is it in agreement with your previous analysis?

```{=plaintext}

```

------------------------------------------------------------------------

# References
